---
title: "STAT535 Final Project"
author: "Marco Lopez"
date: "2024-12-04"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(jsonlite)
library(dplyr)
library(tidyr)
library(rvest)
library(stringr)
library(writexl)
```

## Goal

To compare the mean difference between winter temperatures in Boston vs. Amherst and mean difference between summer temperatures in Boston vs. Amherst to see if one city has a significantly higher/lower average temperature in either season.

Use bootstrapping techniques to compare.

## Data Collection

Collect winter 2024 data from Amherst and Boston in January 2024 and February 2024.

Collect summer 2024 data from Amherst and Boston in June 2024 and July 2024.

We will end up having a winter 2024 dataset for Amherst and Boston as well as a summer 2024 dataset for Amherst and Boston (4 total data sets).

```{r}
# Reading data from excel again::
library(openxlsx)
library(truncnorm)
library(ggplot2)
set.seed(123)
```


```{r}
amherst_summer <- read.xlsx('/Users/malop/Documents/STAT535/Amherst_Summer_2024_Data.xlsx')
x2 <- amherst_summer[,2]
boston_summer <- read.xlsx('/Users/malop/Documents/STAT535/Boston_Summer_2024_Data.xlsx')
xv <- boston_summer[,2]
```

### Amherst Summer

```{r}
# amherst summer
hist(x2,prob=TRUE,breaks = 25)
lines(density(x2),col='blue')
lines(density(rtruncnorm(length(x2),a=40,b=100,mean = mean(x2),sd=sd(x2))),col='red')
lines(density(rgamma(length(x2),shape =mean(x2)^2/sd(x2)^2 ,scale = sd(x2)^2/mean(x2))),col='green')
```

### Boston Summer

```{r}
#boston summer
hist(xv,prob=TRUE,breaks = 20)
lines(density(xv),col='blue')
lines(density(rtruncnorm(length(xv),a=40,b=100,mean = mean(xv),sd=sd(xv))),col='red')
lines(density(rgamma(length(xv),shape =mean(xv)^2/sd(xv)^2 ,scale = sd(xv)^2/mean(xv))),col='green')

```


Simulating data with truncated Normal distribution,
```{r}
#amherst summer

par(mfrow = c(2, 3))
hist(x2, probability = TRUE,  main = "", ylab = "", xlab = "")
for (i in 1:5) {
  x <- rtruncnorm(length(x2),a=40,b=100,mean = mean(x2),sd=sd(x2))
  hist(x, probability = TRUE, col = rgb(1, 0, 0, 0.5),  main = "", ylab = "", xlab = "")
}
```


```{r}

#boston summer
par(mfrow = c(2, 3))
hist(xv, probability = TRUE,  main = "", ylab = "", xlab = "")
for (i in 1:5) {
  x <- rtruncnorm(length(xv),a=40,b=100,mean = mean(xv),sd=sd(xv))
  hist(x, probability = TRUE, col = rgb(1, 0, 0, 0.5),  main = "", ylab = "", xlab = "")
}

```


Power Test:

```{r}
#Setting up parameters functoin:

(boston_tN_par <- 
    boston_summer %>%
    summarise(n = n(), mean = mean(High_Temperature), sd = sd(High_Temperature)) %>%
    mutate(n=n, mean = mean, sd = sd))


(amherst_tN_par <- 
    amherst_summer %>%
    summarise(n = n(), mean = mean(High_Temperature), sd = sd(High_Temperature)) %>%
    mutate(n=n, mean = mean, sd = sd))

amherst_tN_par <- amherst_tN_par %>% mutate(city = "aS")
boston_tN_par <- boston_tN_par %>% mutate(city = "bS")

# joining both datasets into a dataframe.
tN_params <- bind_rows(amherst_tN_par, boston_tN_par)
```

```{r}
true_diff_tN <- mean(x2) - mean(xv)

make_tN_sample_df <- function(tN_params,true_diff){
  
  dat_diff <- tN_params$mean[1] - tN_params$mean[2]
  
  smpl_aS <- rtruncnorm(tN_params$n[1], a= 40, b = 100, mean = tN_params$mean[1], sd=tN_params$sd[1])
  smpl_bS <- rtruncnorm(tN_params$n[2], a= 40, b = 100, mean = tN_params$mean[2], sd=tN_params$sd[2]) + dat_diff + true_diff
  
  return( 
    tibble(
      temperature = c(smpl_aS, smpl_bS),
      city = rep(c("Amherst", "Boston"), tN_params$n)
    )
  ) 
}

```


```{r}

make_tN_sample_df(tN_params, 3) %>%
  ggplot() +
  geom_histogram(aes(x = temperature, fill = city), alpha = 0.5)

make_tN_sample_df(tN_params, 0) %>%
  ggplot() +
  geom_histogram(aes(x = temperature, fill = city), alpha = 0.5)

make_tN_sample_df(tN_params, -3) %>%
  ggplot() +
  geom_histogram(aes(x = temperature, fill = city), alpha = 0.5)

```




```{r}
#t-test
t_test_df <- function(df,alpha = 0.05){
  
  test_result <- t.test(
    x=df$temperature[df$city=='Amherst'],
    y=df$temperature[df$city=='Boston'],
    conf.level = 1-alpha
  )
  
  return(test_result$p.value < alpha)
}

power_estimator <- function(M = 2000, tN_params, true_diff, alpha= 0.05){
  t_reject <- 0
  for (i in 1:M){
    df <- make_tN_sample_df(tN_params,true_diff)
    t_reject <- t_reject + t_test_df(df,alpha = alpha)
  }
  
  return(power_t = t_reject/M)
}

power_estimator(2000, tN_params, -3, 0.05)

power_estimator(2000, tN_params, 0, 0.05)

power_estimator(2000, tN_params, 3, 0.05)

power_estimator(2000, tN_params, -3.31, 0.05)
```


```{r}
#Full power study::
alpha <- 0.05; M <- 500
diff_range <- seq(-6,6,0.5)
power_study <- sapply(diff_range,function(diff){ power_estimator(M=M, tN_params ,true_diff = diff, alpha = alpha) })

```


```{r}
#making the ggplot::
ggplot( data.frame(t_power <- power_study, true_diff = diff_range)
) + 
  
  geom_line(aes(x=true_diff, y=t_power),col='blue') + 
  
  geom_hline(yintercept = alpha, col='red') +
  
  labs(
    x = "True Difference in Means", y = "Statistical Power", title = "Power Curve for t-test" )  +
  
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),  
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  )

```


Trying out gamma:

```{r}

get_gamma_params <- function(mean, sd) {
  a <- mean^2 / sd^2
  s <- sd^2 / mean
  return(list(a = a, s = s))
}


(boston_gamma_par <- 
    boston_summer %>%
    summarise(n = n(), mean = mean(High_Temperature), sd = sd(High_Temperature)) %>%
    mutate(a = get_gamma_params(mean, sd)$a, s = get_gamma_params(mean, sd)$s))


(amherst_gamma_par <- 
    amherst_summer %>%
    summarise(n = n(), mean = mean(High_Temperature), sd = sd(High_Temperature)) %>%
    mutate(a = get_gamma_params(mean, sd)$a, s = get_gamma_params(mean, sd)$s))



```



```{r}
par(mfrow = c(2, 3))
hist(x2, probability = TRUE,  main = "", ylab = "", xlab = "")
for (i in 1:5) {
  x <- rgamma(amherst_gamma_par$n, shape = amherst_gamma_par$a, scale = amherst_gamma_par$s)
  hist(x, probability = TRUE, col = rgb(1, 0, 0, 0.5),  main = "", ylab = "", xlab = "")
}

```


```{r}
par(mfrow = c(2, 3))
hist(xv, probability = TRUE,  main = "", ylab = "", xlab = "")
for (i in 1:5) {
  x <- rgamma(boston_gamma_par$n, shape = boston_gamma_par$a, scale = boston_gamma_par$s)
  hist(x, probability = TRUE, col = rgb(1, 0, 0, 0.5),  main = "", ylab = "", xlab = "")
}
```




```{r}
amherst_gamma_par <- amherst_gamma_par %>% mutate(city = "Amherst")
boston_gamma_par <- boston_gamma_par %>% mutate(city = "Boston")

# Combine the two datasets
gamma_params <- bind_rows(amherst_gamma_par, boston_gamma_par)
gamma_params
```



```{r}


(dat_diff <- gamma_params$mean[1] - gamma_params$mean[2])

gam_params <- gamma_params
make_two_sample_df <- function(gam_params, true_diff) {
  dat_diff <- gam_params$mean[1] - gam_params$mean[2]
  smpl_A <- rgamma(gam_params$n[1], shape = gam_params$a[1], scale = gam_params$s[1])
  smpl_B <- rgamma(gam_params$n[2], shape = gam_params$a[2], scale = gam_params$s[2]) + dat_diff + true_diff
  return( 
    tibble(
      wt = c(smpl_A, smpl_B),
      city = rep(c("Amherst", "Boston"), gam_params$n)
    )
  )
}

# Test generative function 
make_two_sample_df(gamma_params, 3) %>%
  ggplot() +
  geom_histogram(aes(x = wt, fill = city), alpha = 0.5)

make_two_sample_df(gamma_params, 0) %>%
  ggplot() +
  geom_histogram(aes(x = wt, fill = city), alpha = 0.5)

make_two_sample_df(gamma_params, -3) %>%
  ggplot() +
  geom_histogram(aes(x = wt, fill = city), alpha = 0.5)


```



```{r}
power_estimator(1000, gamma_params, 0, 0.05)
power_estimator(2000, gamma_params, -3, 0.05)

# added

power_estimator(2000, tN_params, 3.31, 0.05)
```


```{r}
#Full power study::
alpha <- 0.05; M <- 500
diff_range <- seq(-6,6,0.5)
power_study <- sapply(diff_range,function(diff){ power_estimator(M=M, gamma_params ,true_diff = diff, alpha = alpha) })

```


```{r}
#making the ggplot::
ggplot( data.frame(t_power <- power_study, true_diff = diff_range)
) + 
  
  geom_line(aes(x=true_diff, y=t_power),col='blue') + 
  
  geom_hline(yintercept = alpha, col='red') +
  
  labs(
    x = "True Difference in Means", y = "Statistical Power", title = "Power Curve for t-test" )  +
  
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),  
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  )

```

## Confidence Intervals

```{r}
# Reading data from excel again:
library(openxlsx)
amherst_summer <- read.xlsx('/Users/malop/Documents/STAT535/Amherst_Summer_2024_Data.xlsx')
amherst_winter <- read.xlsx('/Users/malop/Documents/STAT535/Amherst_Winter_2024_Data.xlsx')

```

```{r}
boston_summer <- read.xlsx('/Users/malop/Documents/STAT535/Boston_Summer_2024_Data.xlsx')
boston_winter <- read.xlsx('/Users/malop/Documents/STAT535/Boston_Winter_2024_Data.xlsx')
```

```{r}

n1 <- nrow(amherst_summer)
n2 <- nrow(boston_summer)
(n1==n2)
n <- n1
alpha <- 0.05
M <- 10000

bstp_diff_means <- matrix(0, nrow = M, ncol = 1)

for (m in 1:M) {
  bootstrap_idx <- sample(n, replace = TRUE)
  
  btsp_amherst <- amherst_summer[bootstrap_idx, , drop = FALSE]
  btsp_boston <- boston_summer[bootstrap_idx, , drop = FALSE]
  
  diff_means <- mean(btsp_amherst[, 2]) - mean(btsp_boston[, 2])
  
  bstp_diff_means[m] <- diff_means
}

hist(bstp_diff_means, breaks = 30, main = "Differences in means between Boston and Amherst in Summer",
     xlab = "Difference in Means", col = "lightblue", border = "white")

obs_diff_means <- mean(amherst_summer[, 2]) - mean(boston_summer[, 2])
abline(v = obs_diff_means, col = "red", lwd = 2, lty = 2)

lower_ci <- quantile(bstp_diff_means, 0.025, na.rm = TRUE)
upper_ci <- quantile(bstp_diff_means, 0.975, na.rm = TRUE)

lower_ci
upper_ci

abline(v = lower_ci, col = "blue", lwd = 2, lty = 2)
abline(v = upper_ci, col = "blue", lwd = 2, lty = 2)
```

```{r}

bstp_diff_means <- matrix(0, nrow = M, ncol = 1)

for (m in 1:M) {
  bootstrap_idx <- sample(n, replace = TRUE)

  btsp_amherst <- amherst_winter[bootstrap_idx, , drop = FALSE]
  btsp_boston <- boston_winter[bootstrap_idx, , drop = FALSE]
  
  diff_means <- mean(btsp_amherst[, 2]) - mean(btsp_boston[, 2])
  
  bstp_diff_means[m] <- diff_means
}

hist(bstp_diff_means, breaks = 30, main = "Differences in means between Boston and Amherst in Winter",
     xlab = "Difference in Means", col = "lightblue", border = "white")

obs_diff_means <- mean(amherst_winter[, 2]) - mean(boston_winter[, 2])
abline(v = obs_diff_means, col = "red", lwd = 2, lty = 2)

# Add confidence intervals

lower_ci <- quantile(bstp_diff_means, 0.025, na.rm = TRUE)
upper_ci <- quantile(bstp_diff_means, 0.975, na.rm = TRUE)
lower_ci
upper_ci

abline(v = lower_ci, col = "blue", lwd = 2, lty = 2)
abline(v = upper_ci, col = "blue", lwd = 2, lty = 2)

```

## Conclusion

We find that there is a statistically significant difference in means between Amherst and Boston in the summer months. There is not a statistically significant difference in means in the winter months.
